Modelling state-switching functional data with hidden Markov models

### 2020-11-15 Report/Sections/intro.tex
small updates

Biologging technology now provides researchers with data collected almost continuously in time \citep{Hooten:2017}.
The collection and analysis of data from devices such as accelerometers has brought new insights to areas ranging from monitoring machine health \citep{Getman:2009} to understanding physical activity levels in children \citep{Morris:2007}. In particular, the study of animal movement has been transformed by advanced tracking devices which collect biologging data \citep{Borger:2020}. Tags can record observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours. These data contain a wealth of information about human and animal behaviour, but parsing these large data sets poses a new challenge for statisticians.

Biologging data are sometimes modelled as a set of curves analyzed by methods of Functional Data Analysis (FDA, see, e.g., \citet{Ramsay:2005}). For example, \citet{Morris:2007} view a child's activity level throughout the day as a curve where metabolic activity is a function of time. Similarly, \citet{Fu:2017} view the dive profile of elephant seal (\textit{Mirounga leonina}) as a sequence of dive curves whereby the amplitude and phase variation of the dive is used to classify dive types.
%, although they ignore sequential dependence between dives. 
%Similarly, our case study views the dive profile of a killer whale (\textit{Orcinus orca}) as a sequence of curves where three orthogonal axes of acceleration are each functions of time.
%
FDA was originally developed to process curves assumed to be independent replicates, but time series data of this type are often complicated by dependencies which violate this assumption. In particular, biologging data often display significant temporal dependence both between curves (coarse-scale) and within curves (fine-scale) \citep{Barajas:2017}.
%, as is the case for our case study data. Namely, the dive profile (see Figure \ref{fig:data}) of a killer whale from the threatened northern resident population off the coast of British Columbia, Canada, clearly shows temporal dependence between dives, as dives of various time durations and depths appear to be clustered together in time. 
For example, marine mammal dive profiles show distinct dive ``types" that appear to cluster together in time  \citep{Tennessen:2019b} while simultaneously displaying bouts of short-term periodicity within dives \citep{Adam:2019}. Fine-scale periodicity within a larger process in particular is common in fields ranging from machine health \citep{Xin:2018,Lucero:2019} to speech recognition \citep{Juang:1991}.

>>>>

Biologging technology now provides researchers with data collected almost continuously in time \citep{Hooten:2017}.
The collection and analysis of data from devices such as accelerometers has brought new insights to areas ranging from monitoring machine health \citep{Getman:2009} to understanding physical activity levels in children \citep{Morris:2007}. In particular, the study of animal movement has been transformed by advanced tracking devices which collect biologging data \citep{Borger:2020}. Tags can record observations at over 50 times per second, resulting in time series containing millions of observations over the course of several hours. These data contain a wealth of information about human and animal behaviour, but parsing these large data sets poses a new challenge for statisticians.

Biologging data are frequently modelled as a set of curves analyzed by methods of Functional Data Analysis (FDA, see, e.g., \citet{Ramsay:2005}). For example, \citet{Morris:2007} view a child's activity level throughout the day as a curve where metabolic activity is a function of time. Similarly, \citet{Fu:2017} view the dive profile of elephant seal (\textit{Mirounga leonina}) as a sequence of dive curves whereby the amplitude and phase variation of the dive is used to classify dive types.
%, although they ignore sequential dependence between dives. 
%Similarly, our case study views the dive profile of a killer whale (\textit{Orcinus orca}) as a sequence of curves where three orthogonal axes of acceleration are each functions of time.
%
FDA was originally developed to process curves assumed to be independent replicates, but time series data of this type are often complicated by dependencies which violate this assumption. In particular, biologging data often display significant temporal dependence both between curves (coarse-scale) and within curves (fine-scale) \citep{Barajas:2017}.
%, as is the case for our case study data. Namely, the dive profile (see Figure \ref{fig:data}) of a killer whale from the threatened northern resident population off the coast of British Columbia, Canada, clearly shows temporal dependence between dives, as dives of various time durations and depths appear to be clustered together in time. 
For example, marine mammal dive profiles show distinct dive ``types" that appear to cluster together in time  \citep{Tennessen:2019b} while simultaneously displaying bouts of short-term periodicity within dives \citep{Adam:2019}. Fine-scale periodicity within a larger process in particular is also common in fields ranging from machine health \citep{Xin:2018,Lucero:2019} to speech recognition \citep{Juang:1991}.

### 2020-11-15 Report/Sections/models.tex
small updates

Hidden Markov models describe state-switching Markovian processes in discrete time and will be the core structure of the models we use for $Y$ and $Y^*$. We first detail the structure of a traditional HMM. We then review a variation described by \cite{Lawler:2019} called the conditionally auto-regressive hidden Markov model (CarHMM), which explicitly models auto-correlation between observations. Either the HMM or the CarHMM can be used to model a sequence of observations at one of the two scales: for the coarse scale sequence $Y$ or for each of the fine scale sequences, $Y_1^*,\ldots, Y_T^*$. To simultaneously model $Y$ and $Y^*$, we use the hierarchical hidden Markov model (HHMM) \citep{Barajas:2017,Adam:2019}.
%which is comprised of two distinct,x but related, HMMs that jointly model these time series. 
Next, we describe how to reduce fine-scale correlation by transforming raw observations into meaningful summary statistics. In particular, we apply a discrete Fourier transform (DFT) to the components of $Y^*$ using a moving window to summarize its periodic behaviour. Finally, we show how each of these models can be altered and combined to form a wide variety of new models.

>>>>

Hidden Markov models describe state-switching Markovian processes in discrete time and will be the core structure of the models we use for $Y$ and $Y^*$. We first detail the structure of a traditional HMM and a variation described by \cite{Lawler:2019} called the conditionally auto-regressive hidden Markov model (CarHMM) which explicitly models auto-correlation between observations. Either the HMM or the CarHMM can be used to model a sequence of observations at one of the two scales: for the coarse scale sequence $Y$ or for each of the fine scale sequences, $Y_1^*,\ldots, Y_T^*$. To simultaneously model $Y$ and $Y^*$, we use the hierarchical hidden Markov model (HHMM) \citep{Barajas:2017,Adam:2019}.
%which is comprised of two distinct,x but related, HMMs that jointly model these time series. 
Next, we describe how to reduce fine-scale correlation by transforming raw observations into meaningful summary statistics. In particular, we apply a discrete Fourier transform (DFT) to the components of $Y^*$ using a moving window to summarize its periodic behaviour. Finally, we show how each of these models can be altered and combined to form a wide variety of new models.

### 2020-11-15 Report/main.tex
small updates

Sequences of curves sampled at high-frequency in time are increasingly common in practice, but they can exhibit complicated dependence structures which cannot be modelled using common methods of Functional Data Analysis (FDA). We detail a hierarchical approach which treats the curves as observations from a hidden Markov model (HMM). The distribution of each curve is then defined by another fine-scale model which may involve a variety of methods such as moving-window transformations, Fourier analysis, and auto-regression. As a case study, we use this framework to model the movement of a killer whale off the western coast of Canada. Through simulations we show that our model produces more interpretable state estimation and more accurate parameter estimates compared to existing baselines. This approach is broadly applicable to sequences of curves exhibiting intricate dependence structures.

>>>>

Data sets comprised of sequences of curves sampled at high-frequency in time are increasingly common in practice, but they can exhibit complicated dependence structures which cannot be modelled using common methods of Functional Data Analysis (FDA). We detail a hierarchical approach which treats the curves as observations from a hidden Markov model (HMM). The distribution of each curve is then defined by another fine-scale model which may involve a variety of methods such as moving-window transformations, Fourier analysis, and auto-regression. This approach is broadly applicable to sequences of curves exhibiting intricate dependence structures. As a case study, we use this framework to model the movement of a killer whale off the western coast of Canada. Through simulations we show that our model produces more interpretable state estimation and more accurate parameter estimates compared to existing baselines.
